---
title: 'Lab 3 & 4: Vector Operations'
author: "Sophia Leiker"
date: "10/12/2022"
output: html_document
---
# introduction

In this lab, we’ll explore the basics of spatial and geometry operations on vector data in R using the sf package. The following materials are modified from Chapter 4 and Chapter 5of Geocomputation with R by Rovin Lovelace.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(sf)
library(spData)
library(tmap)
library(tidyverse)
library(rmapshaper)
library(smoothr)
```

# Vector Operations

```{r}
# Looking at initial datasets to be used
nz
nz_height
```

## Spatial Subsetting

Spatial subsetting is the process of converting a spatial object into a new object containing only the features that relate in space to another object. This is analogous the attribute subsetting that we covered last week. There are many ways to spatially subset in R, so we will explore a few.

Let’s start by going back to the New Zealand datasets and find all the high points in the state of Canterbury.

```{r}
#filtering to only select canterbury
canterbury <- nz %>% 
  filter(Name == "Canterbury")

#this gives us all the points that fall inside of canterbury
c_height <- nz_height[canterbury,]

#making a map to check
tm_shape(nz) +
  tm_polygons() +
  tm_shape(nz_height)+
  tm_dots(col = "red")

#only mapping the points that fall into canterbury
tm_shape(nz) +
  tm_polygons() +
  tm_shape(c_height) +
  tm_dots(col = "red")

# The default is to subset to features that intersect, but we can use other operations, including finding features that do not intersect.
#creating for opposite, using st_dis
outside_height <- nz_height[canterbury, , op = st_disjoint]

#map it
tm_shape(nz) +
  tm_polygons() +
  tm_shape(outside_height) +
  tm_dots(col = "red")

#We can perform the same operations using topological operators. These operators return matrices testing the relationship between features.
#st_intersects (if it touches at all, include it)

sel_sgbp <- st_intersects(x = nz_height, y = canterbury)

#turning this into a logical (anything greater than 0 are the rows we want because that means they intersect it)
sel_logical <- lengths(sel_sgbp) > 0

c_height2 <-nz_height[sel_logical, ]
tm_shape(nz) +
  tm_polygons() +
  tm_shape(c_height2) +
  tm_dots(col = "red")
```

Points in Canterbury Again
We can also use the `st_filter` function in sf

```{r}
#Can do the same thing with tidy commands using st_filter
c_height3 <- nz_height %>% 
  st_filter(y=canterbury, .predicate = st_intersects) #filter using a spatial intersection

tm_shape(nz) +
  tm_polygons() +
  tm_shape(c_height3) +
  tm_dots(col = "red")
```
We can change the predicate option to test subset to features that don’t intersect

```{r}
#We can change the predicate option to test subset to features that don’t intersect
outside_height10 <- nz_height %>% 
  st_filter(y=canterbury, .predicate = st_disjoint)

tm_shape(nz) +
  tm_polygons() +
  tm_shape(outside_height10) +
  tm_dots(col = "red")
```

## Creating Bounding Box

## Spatial joining

Where attribute joining depends on both data sets sharing a ‘key’ variable, spatial joining uses the same concept but depends on spatial relationships between data sets.

Let’s test this out by creating 50 points randomly distributed across the world and finding out what countries they call in.

```{r}
#creating bounding box

bb <- st_bbox(world)

#setting random points within the dataframe
random_df <- data.frame(
  x=runif(n=10,min = bb[1], max = bb[3]),
  y=runif(n=10,min = bb[2], max = bb[4])
)

#turning a data frame into an sf object
random_points <- random_df %>% 
  st_as_sf(coords = c("x", "y")) %>% 
  st_set_crs("EPSG:4326")

#checking that random points is now an sf object
class(random_points)

#mapping this
tm_shape(world) +
  tm_fill() +
  tm_shape(random_points) +
  tm_dots(col = "red")

#figuring out what countries have the points in them
world_random <- world[random_points, ]
tm_shape(world_random) +
  tm_fill() #just filling in shapes that have points in them
```

```{r}
#Practicing Using World Random again 
#Let’s first use spatial subsetting to find just the countries that contain random points.
bb <- st_bbox(world)

random_df <- data.frame(
  x = runif(n = 10, min = bb[1], max = bb[3]), #bounding box for the x is 1 and 3
  y = runif(n=10, min = bb[2], max = bb[4]) #bounding box for the y is 2 and 4
) #this will create a datframe with 10 values with x and y objects 


world_random <- world[random_points,] #this is how to spatially subset using the rows which align in each data frame provided

random_points <- random_df %>%
  st_as_sf(coords = c("x", "y")) %>% #c with the name of the x and y values
  st_set_crs("EPSG:4326") #setting the coordinate reference system 

tm_shape(world) +
  tm_fill() +
  tm_shape(random_points) +
  tm_dots(col = "red")

#Now let’s perform a spatial join to add the info from each country that a point falls into onto the point dataset.
random_joined <- st_join(random_points, world) #joining points that fall inside the countries 

#By default, st_join performs a left join. We change this and instead perform an inner join.
random_joined_inner <- st_join(random_points, world, left = FALSE) #joining points that fall outside of the countries
```

## Non-Overlapping Joins
Sometimes we might want join geographic datasets that are strongly related, but do not have overlapping geometries. To demonstrate this, let’s look at data on cycle hire points in London.

```{r}
#Non-overlapping joins 

tmap_mode("view")

tm_shape(cycle_hire) +
  tm_dots(col = "blue", alpha = 0.5) + #alpha is changing the transparency 
  tm_shape(cycle_hire_osm) +
  tm_dots(col = "red", alpha = 0.5)

#We can check if any of these points overlap.
any(st_touches(cycle_hire, cycle_hire_osm, sparse = FALSE)) #none of these points actually touch

#Let’s say we need to join the ‘capacity’ variable in ‘cycle_hire_osm’ onto the official ‘target’ data in ‘cycle_hire’. The simplest method is using the topological operator st_is_within_distance().

sel <- st_is_within_distance(cycle_hire, cycle_hire_osm, dist = 20)
summary(lengths(sel) > 0) #summarizes the number of points within 20 meters

#Now, we’d like to add the values from ‘cycle_hire_osm’ onto the ‘cycle_hire’ points.
z <- st_join(cycle_hire, cycle_hire_osm, st_is_within_distance, dist = 20)
nrow(cycle_hire) #742 rows in original data
nrow(z) #762 rows in new data

#Note: the number of rows of the join is larger than the number of rows in the original dataset. Why? Because some points in ‘cycle_hire’ were within 20 meters of multiple points in ‘cycle_hire_osm’. If we wanted to aggregate so we have just one value per original point, we can use the aggregation methods from last week.
z <- z %>% 
  group_by(id) %>% 
  summarize(capacity = mean(capacity))

nrow(z) #now they are the same length

```

## Spatial Aggregation

Similar to attribute data aggregation, spatial data aggregation condenses data (we end up with fewer rows than we started with).

Let’s say we wanted to find the average height of high point in each region of New Zealand. We could use the aggregate function in base R.
```{r}
nz_agg <- aggregate(x = nz_height, by = nz, FUN = mean)
#x object is the information we want to aggregate, and we want to aggregate it by the y value, then the FUN indicates the function we want to use to aggregate 
```

The result of this is an object with the same geometries as the aggregating feature data set (in this case ‘nz’).

We could also use a `sf/dplyr` approach.

```{r}
nz_agg <- st_join(nz, nz_height) %>% #joining onto the aggregating dataset so nz is on the left (because that is what we are joining onto)
  group_by(Name) %>% 
  summarize(elevation = mean(elevation, na.rm = TRUE))
```

## joining incongruent layers

We might want to aggregate data to geometries that are not congruent (i.e. their boundaries don’t line up). This causes issues when we think about how to summarize associated values.

```{r}
head(incongruent)
head(aggregating_zones)
```

```{r}
tm_shape(incongruent) +
  tm_polygons() +
  tm_shape(aggregating_zones) +
  tm_borders(col = "red")
```

The simplest method for dealing with this is using area weighted spatial interpolation which transfers values from the ‘incongruent’ object to a new column in ‘aggregating_zones’ in proportion with the area of overlap.

```{r}
iv <- incongruent["value"]
agg_aw <- st_interpolate_aw(iv, aggregating_zones, extensive = TRUE) #aggregating zones are what we are aggregating to

tm_shape(agg_aw) +
  tm_fill(col = "value")
```

## distance relationships

While topological relationships are binary (features either intersect or don’t), distance relationships are continuous.

We can use the following to find the distance between the highest point in NZ and the centroid of the Canterbury region.

```{r}
nz_highest <- nz_height %>% 
  slice_max(n = 1, order_by = elevation) #this will give us the highest point in new zealand


canterbury <- nz %>% 
  filter(Name == "Canterbury")

#finding the centroid of canterbury
canterbury_centroid = st_centroid(canterbury)

#finding the distance
st_distance(nz_highest, canterbury_centroid)

#Note: this function returns distances with units (yay!) and as a matrix, meaning we could find the distance between many locations at once.
```

## Simplification

Simplification generalizes vector data (polygons and lines) to assist with plotting and reducing the amount of memory, disk space, and network bandwidth to handle a dataset.

Let’s try simplifying the US states using the Douglas-Peucker algorithm. GEOS assumes a projected CRS, so we first need to project the data, in this case into the US National Atlas Equal Area (epsg = 2163)

```{r}
us_states2163 <- st_transform(us_states, "EPSG:2163")
us_states2163 = us_states2163

us_states_simp1 = st_simplify(us_states2163, dTolerance = 100000)

tm_shape(us_states_simp1) +
  tm_polygons()


```

```{r}
# proportion of points to retain (0-1; default 0.05)
us_states_simp2 = rmapshaper::ms_simplify(us_states2163, keep = 0.01, keep_shapes = TRUE) #this keeps the shapes/connections between the states

#mapping this

tm_shape(us_states_simp2) +
  tm_polygons()
```


Instead of simplifying, we could try smoothing using Gaussian kernel regression.
```{r}
us_states_simp3 = smoothr::smooth(us_states2163, method = 'ksmooth', smoothness = 6)

tm_shape(us_states_simp3) +
  tm_polygons()
```

## centroids
Centroids identify the center of a spatial feature. Similar to taking an average, there are many ways to compute a centroid. The most common is the geographic centroid.

```{r}
nz_centroid <- st_centroid(nz)

tm_shape(nz) +
  tm_fill() +
  tm_shape(nz_centroid) +
  tm_dots()

```

Sometimes centroids fall outside of the boundaries of the objects they were created from. In the case where we need them to fall inside of the feature, we can use point on surface methods.

```{r}
nz_pos <- st_point_on_surface(nz)

tm_shape(nz) +
  tm_fill() +
  tm_shape(nz_centroid) +
  tm_dots() +
  tm_shape(nz_pos) +
  tm_dots(col = "red")
```

## buffers
Buffers create polygons representing a set distance from a feature.

```{r}
seine_buffer <- st_buffer(seine, dist = 5000)

tm_shape(seine_buffer) +
  tm_polygons()
```

## unions
As we saw in the last lab, we can spatially aggregate without explicitly asking R to do so.


```{r}
w <- tm_shape(world) +
  tm_polygons()

continents <- world %>%
  group_by(continent) %>%
  summarize(population = sum(pop, na.rm = TRUE))

c <- tm_shape(continents) +
  tm_polygons()

tmap_arrange(w, c)

```

What is going on here? Behind the scenes, summarize() is using st_union() to dissolve the boundaries.
```{r}
us_west <- us_states %>%
  filter(REGION == "West")

us_west_union <- st_union(us_west)

tm_shape(us_west_union) +
  tm_polygons()
```

st_union() can also take 2 geometries and unite them.

```{r}
texas <-  us_states %>%
  filter(NAME == "Texas")

texas_union = st_union(us_west_union, texas)

tm_shape(texas_union) +
  tm_polygons()
```

